\begin{acronym}[OME-TIFF]
    \acro{ABS}[ABS]{Acrylonitrile butadiene styrene, a printable thermoplastic used for stronger prototype parts.}
    \acro{ABSGF}[ABS-GF]{Glass-fiber-reinforced acrylonitrile butadiene styrene, used for stiffer printed parts.}
    \acro{ADC}[ADC]{Analog-to-digital converter, used to convert analog joystick voltages into digital values.}
    \acro{AF}[AF]{Autofocus.}
    \acro{AI}[AI]{Artificial intelligence.}
    \acro{API}[API]{Application programming interface, a software interface exposed for other programs.}
    \acro{BCM}[BCM]{Broadcom GPIO numbering scheme used by the Raspberry Pi.}
    \acro{BSS138}[BSS138]{A MOSFET commonly used in small logic-level shifter modules.}
    \acro{CAD}[CAD]{Computer-aided design.}
    \acro{CPU}[CPU]{Central processing unit.}
    \acro{CSV}[CSV]{Comma-separated values, a plain-text table format used for evaluation results.}
    \acro{DoF}[DoF]{Depth of field.}
    \acro{DSI}[DSI]{Display Serial Interface, used by the Raspberry Pi touchscreen.}
    \acro{FIFO}[FIFO]{First in, first out, a queue behavior used in serial communication buffers.}
    \acro{GB}[GB]{Gigabyte.}
    \acro{GND}[GND]{Ground reference in the electronics wiring.}
    \acro{GPIO}[GPIO]{General-purpose input/output pins on the Raspberry Pi.}
    \acro{I2C}[I\textsuperscript{2}C]{Inter-Integrated Circuit, a serial bus used by the ADC, current sensors, and touchscreen controller.}
    \acro{JSON}[JSON]{JavaScript Object Notation, a structured text format used for configuration and API data.}
    \acro{JST}[JST]{A family of small electrical connector types used in wiring harnesses.}
    \acro{NA}[NA]{Numerical aperture of a microscope objective.}
    \acro{OME}[OME]{Open Microscopy Environment.}
    \acro{OMETIFF}[OME-TIFF]{Open Microscopy Environment TIFF, a microscopy image format with structured metadata.}
    \acro{OMEXML}[OME-XML]{Open Microscopy Environment XML metadata embedded in or associated with microscopy image files.}
    \acro{OpenCV}[OpenCV]{Open Source Computer Vision Library.}
    \acro{PCB}[PCB]{Printed circuit board.}
    \acro{PETG}[PETG]{Polyethylene terephthalate glycol, a 3D-printing filament material.}
    \acro{PNG}[PNG]{Portable Network Graphics image format.}
    \acro{QML}[QML]{Qt Modeling Language, the declarative language used for the touchscreen user interface.}
    \acro{REST}[REST]{Representational State Transfer, an architectural style used for the HTTP API.}
    \acro{RQ}[RQ]{Research question.}
    \acro{SCL}[SCL]{Serial clock line of an I\textsuperscript{2}C bus.}
    \acro{SDA}[SDA]{Serial data line of an I\textsuperscript{2}C bus.}
    \acro{STEP}[STEP]{Standard for the Exchange of Product Data, a CAD exchange file format.}
    \acro{TIFF}[TIFF]{Tagged Image File Format.}
    \acro{UART}[UART]{Universal asynchronous receiver-transmitter, a serial communication interface.}
    \acro{UI}[UI]{User interface.}
    \acro{USB}[USB]{Universal Serial Bus.}
    \acro{UVC}[UVC]{USB Video Class, a standard for USB camera and capture devices.}
    \acro{V4L2}[V4L2]{Video4Linux2, the Linux video capture interface.}
\end{acronym}
